Sequential rank agreement methods for comparison of ranked lists

### 2016-03-31 latex/paper.tex
Introducing population-level definition

\documentclass[12pt,a4paper]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\usepackage{subcaption}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage{amsmath}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
%\SweaveOpts{concordance=TRUE}

\title{Sequential rank agreement methods for comparison of ranked lists}
\author{}

@@

\begin{abstract}

\end{abstract}

@@

Ranked lists occur in many applications of statistics. Regression
methods rank predictor variables according to magnitude of association
with outcome, prediction models rank subjects according to their risk
of an event, and genetic studies rank genes according to their
difference in expression across samples. A common research question is
where to stop, i.e., to decide the maximal significant rank.  In the
three examples this would correspond to the number of significantly
associated predictor variables, the number of patients at high risk of
an event, and the number of genes that are worth to pursue in further
experiments, respectively.

In this article we describe some new breakthrough tools for measuring
agreement across a set of lists which soon will enter the state-of-the
art. The methods should be useful whenever there are multiple rankings
of the same list. The idea is to define agreement based on the ranks
of the first k elements in each list.

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$. An
ordered list is a permutation function, $R: \{1,\dots,P\}\to
\{1,\dots,P\}$, such that $R(X_p)$ is the rank of item $X_p$ in the
list. The inverse mapping $\pi=R^{-1}$ assigns to rank
$r\in\{1,\dots,P\}$ the item $\pi(r)$ found at that rank. The methods
described below work for a set of $L$ lists $R_1,\dots,R_L$,
$L\geq2$. We denote $\pi_l=R_l^{-1}$ for the corresponding inverse
mappings. Panels (a) and (b) of Table~\ref{tab:example} show a
schematic example of these mappings.

\begin{table}[tb]
  \caption{Example set of ranked lists. (a) shows the ranked list of items for each of three lists, (b) presents the ranks obtained by each item in each of the three lists and (c) shows the cumulative set of items up to a given depth in the three lists.}
\begin{center}
  \begin{subtable}{4cm}%
    \caption{}
      \begin{tabular}{cccc}
        \hline\hline % & \multicolumn{3}{c}{List} \\
        Rank & $\pi_1$ & $\pi_2$ & $\pi_3$ \\ \hline
        1 & A & A & B \\
        2 & B & C & A \\
        3 & C & D & E \\
        4 & D & B & C \\
        5 & E & E & D \\ \hline
    \end{tabular}
  \end{subtable}
% \hfill
\hspace{1em}
%%% 
  \begin{subtable}{4cm}%
    \caption{}
    \begin{tabular}{cccc} \hline\hline
  % & \multicolumn{3}{c}{List} \\
    Item & $R_1$ & $R_2$ & $R_3$ \\ \hline
    A & 1 & 1 & 2 \\
    B & 2 & 4 & 1 \\
    C & 3 & 2 & 4 \\
    D & 4 & 3 & 5 \\
    E & 5 & 5 & 3 \\ \hline
  \end{tabular}
\end{subtable}
% \hfill
\hspace{1em}
%%%
\begin{subtable}{4cm}%
  \caption{}
\begin{tabular}{cc} \hline\hline
% Depth   &     \\
Depth  &  $S_d$ \\ \hline
1 & $\{$A, B$\}$\\
2 & $\{$A, B, C$\}$ \\
3 & $\{$A, B, C, D, E$\}$ \\
4 & $\{$A, B, C, D, E$\}$ \\
5 & $\{$A, B, C, D, E$\}$ \\ \hline
\end{tabular}
\end{subtable}
\end{center}
\label{tab:example}
\end{table}

The agreement of the lists regarding the rank given to an item can be
measured by
\begin{equation}
  A(p) = f(R_1(p), \ldots, R_L(p)),
\end{equation}
for a distance function $f$. Throughout this paper we will use the
sample standard error as our function $f$ and hence use
$$A(p) = \sqrt{\frac{\sum_{i=1}^L (R_i(p) - \bar{R}(p))^2}{L-1}},
$$
but other choices could be made (see the discussion). The sample
standard error has an interpretation as the average distance of the
individual rankings of the lists from the average ranking.

We now describe what is exemplified in Panel (c) of Table
\ref{tab:example} and how it can be used to define \emph{sequential
  rank agreement}. For an integer $1\le d\le P$ we define the unique
set of items found in the $L$ top $d$ parts of the lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{\pi_l(r) ; r \leq d, l = 1, \ldots, L \}.
\end{equation}
The \emph{sequential rank agreement} is the pooled standard deviation
of the items found in the set $S_d$:
\begin{equation}
\textrm{SRA}(d)= \sqrt{\frac{\sum_{\{p \in S_d\}}(L-1)A(p)^2}{(L-1)|S_d|}}.
\end{equation}

\subsection{All lists fully observed}
We shall start by the simplest case where all $L$ lists are fully observed.

\subsection{Analysis of top $k$ lists}
Not uncommon for lists to be

censored

Let $\Lambda_l, l=1, \ldots, L$ be the set of items found in list $l$
so $\Lambda_l$ is the top $k_l$ list of items from list $l$ where $k_l
= |\Lambda_l|$. Note that we observe the top $k$ items for each of the
$L$ lists if $k_1=\cdots=k_L=k$. For censored lists the rank function
becomes
\begin{equation}
\tilde R_l(p) = \left\{\begin{array}{cl} \{\pi_l^{-1}(p)\} & \text{ for } p\in \Lambda_l \\ 
\{k_l+1,\dots,P\} & \text{ for } p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$ must be larger than the largest rank observed in that list.

In the case of censored lists it is sufficient (FIXME: requires
argument) to look at depths where we have corresponding observations
so the largest rank we should consider will be
\begin{equation}
d \leq \max(k_1, \ldots, k_L).
\end{equation}

We cannot directly compute $A(p)$ for all predictors because we only observe a censored version of $\tilde R$ for some of the lists. Instead we assume that the rank assigned to predictor $p$ in list $l$ is uniformly distributed among the ranks that have been unassigned for list $l$. The ranks are clearly not independent since each of the lists essentially contains full set of ranks

\begin{equation}
\tilde A(p) = \frac{\sum_{r_1; r_1\in \tilde R_1(p)}  \cdots \sum_{r_L; r_L\in \tilde R_L(p)} A(p)}{\prod_l |\tilde R_l(p)|}
\end{equation}

FIXME: If instead of running through all elements of $\tilde R_1(p)$
one would use the average rank in $\tilde R_1(p)$ we would end up with
a too small variance.

\section{Benchmarks}

\section{Applications}

\subsection{Evaluating results from top-$k$ lists}

\subsection{Comparing results across different method}

\subsection{Stability of results}

Bootstrap across a single method and compare results. Discuss collinearity

\section{Discussion}

Mention/discuss different measures.

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[]{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[]{color}
\usepackage{subcaption}
\usepackage{natbib}
\usepackage{xspace}
\usepackage{calc}
\usepackage[noae]{Sweave}
\usepackage{authblk}
\usepackage{algorithm}
\usepackage{algpseudocode}
%\usepackage{geometry}
%\geometry{a4paper, portrait, margin=1in}
%
% New commands
%
\newcommand{\ie}{\emph{i.e.}\@\xspace}
\newcommand{\eg}{\emph{e.g.}\@\xspace}

\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}

\usepackage[amsthm,thmmarks]{ntheorem}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem*{proposition2}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{assumption}{Assumption}
\newtheorem{example}{Example}
\renewcommand\proofSymbol{\ensuremath{\blacksquare}}

%%% RSS specific items

\title{Sequential rank agreement methods for comparison of ranked lists}
\author{Claus Thorn Ekstrøm}
\author{Thomas Alexander Gerds}
\author{Andreas Kryger Jensen}
\author{Kasper Brink-Jensen}
\affil{Biostatistics, University of Copenhagen}
\renewcommand{\topfraction}{.85}
\renewcommand{\bottomfraction}{.7}
\renewcommand{\textfraction}{.15}
\renewcommand{\floatpagefraction}{.66}
\renewcommand{\dbltopfraction}{.66}
\renewcommand{\dblfloatpagefraction}{.66}
\setcounter{topnumber}{9}
\setcounter{bottomnumber}{9}
\setcounter{totalnumber}{20}
\setcounter{dbltopnumber}{9}

\begin{document}
\input{paper-concordance}

% Use square brackets for numbering affiliations

@@

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and prominent task in applied statistics. Predictor
  variables are ranked according to magnitude of association with an
  outcome, prediction models rank subjects according to the
  personalized risk of an event, and genetic studies rank genes
  according to their difference in gene expression levels. This
  article constructs measures of the agreement of two or more ordered
  lists. We use the standard deviation of the ranks to define a
  measure of agreement that both provides an intuitive interpretation
  and can be applied to any number of lists even if some or all are
  incomplete or censored. The approach can identify change-points in
  the agreement of the lists and the sequential changes of agreement
  as a function of the depth of the lists can be compared graphically
  to a permutation based reference set.  The usefulness of these tools
  are illustrated using gene rankings, and using data from two Danish
  ovarian cancer studies where we assess the within and between
  agreement of different statistical classification methods.
\end{abstract}

Key words: sequential rank agreement, partial ranking, order
statistic, gene rankings, methods comparison, variable selection,
permutation

@@

Ranking of items or results is common in scientific research and
ranked lists occur naturally as the result of many statistical
applications. Regression methods rank predictor variables according to
magnitude of their association with an outcome, prediction models rank
subjects according to their risk of an event, and genetic studies rank
genes according to their difference in gene expression levels across
samples.

When several rankings of the same items are available, a common
research question is to what extent they agree. In particular, is it
possible to identify an optimal rank until which the lists agree on
the items?  A typical situation arises in high-dimensional genomics
studies when several analysis methods are applied to rank a list of
genes according to their association with a phenotype, treatment
effect or other outcome. A measure of agreement of gene rankings
obtained by different methods could often help to identify which genes
are worth to pursue in further experiments.

Several approaches exist to measure the ``distance'' between two
ranked lists. Of these, Kendall's $\tau$ \citep{kendall1948} and
Spearman’s footrule $\rho$ \citep{Spearman1910} are among the most
well-known, but these measures do not distinguish between agreement in
the top versus towards the bottom of the lists and the provide a
single measure of the overall distance. \citet{Shieh1998} proposed a
weighted version of Kendall's $\tau$ where each pair of rankings can
be assigned different weights, and \citet{Yilmaz2008} proposed the
$\tau_{ap}$ which places higher emphasis on the top of the lists.
Spearman’s footrule uses the ranks of the variables for calculation of
the distance and the use of ranks are also employed in the $M$ measure
of \citet{Bar-Ilan2006} where the reciprocal rank differences are used
to calculate the similarity measure.

Other recent approaches consider the intersection of lists as the basis
for a similarity measure. However, simple intersection also places
equal weights on all depths of the list and therefore \citet{Fagin2003} and
\citet{Webber2010} proposed weighted intersections which put more
emphasis on the top of the lists. One proposal is denoted the average
overlap (AO) and used for comparison later on. Specifically,
\citet{Webber2010} define their rank-biased overlap (RBO) by weighting
with a converging series to ensure that the top is weighted higher
than the potentially non-informative bottom of the lists.
It is possible to use the existing methods to calculate agreement of
lists until a given depth, i.e., limited to the $d$ items of each
list. However, the interpretation may not be straightforward,
especially in the case of more than two lists, and they may not
accommodate partial rankings.

In this article we introduce sequential rank agreement for measuring
agreement among (partially) ranked lists.  The general idea is to
define agreement based on the sequence of ranks from the first $d$
elements in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{alt:bland:1983,Carstensen2010}. Our proposed approach allows us
to compare more than two lists simultaneously, it provides a dynamic
measure of agreement as a function of the depth in the lists, places
high weight on the top of the list, accommodates censored/incomplete
lists of varying lengths, and has a natural interpretation that
directly relates to the ranks. Graphical illustration of sequential
rank agreement potentially allows us to infer a change-point, i.e., a
list depth where a change in the agreement of the lists occur but we
also provide randomization-based graphical tools to compare the
observed rank agreement to the expected agreement found in
non-informative data.  In this sense our approach is a combination and
generalization of some of the ideas of \citet{Carterette2009} and
\citet{Boulesteix2009}. The former compares two rankings based on the
distance between them as measured by a multivariate Gaussian
distribution and the latter presents an overview of approaches for
aggregation of ranked lists including bootstrap and leave-one-out
jackknife approaches.

The manuscript is organized as follows: In the next section we define
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete/censored/partial lists. In section 3 we present and
discuss approaches to evaluate the results obtained from sequential
rank agreement. Finally we apply the sequential rank agreement to two
Danish ovarian cancer studies before we discuss the findings along
with possible extensions. The approaches presented in this manuscript
are available in the R package \texttt{SuperRanker}.

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$ and a
ranking function $R: \{X_1,\dots,X_P\}\to \{1,\dots,P\}$, such that
$R(X_p)$ is the rank of item $X_p$. The inverse mapping $R^{-1}$ gives
the item $R^{-1}(r)$ that was assigned to rank $r\in\{1,\dots,P\}$. An
ordered list is the realization of a ranking function $R$ applied to a
data sample $X$ and we let $R_1(X),\dots,R_L(X)$, $L\geq2$, denote
ordered lists obtained from $L$ different ranking functions applied to
the same data $X$. We will use the same notation for the equivalent
scenario where we have ordered lists obtained from applying the same
ranking function to $L$ samples from the same population, i.e., for
ranked lists $R(X^1), \ldots, R(X^L)$.  Panels (a) and (b) of
Table~\ref{tab:example} show a schematic example of these
mappings. Thus if $R_l^{-1}(1)=X_{34}$ then item $X_{34}$ is ranked
first in list $l$ and similarly $R_l(X_{34})=1$.

\begin{table}[tb]
  \caption{Example set of ranked lists. (a) shows the ranked list of
    items for each of three lists, (b) presents the ranks obtained by
    each item in each of the three lists and (c) shows the cumulative
    set of items up to a given depth in the three lists.}
\begin{center}
\footnotesize{
  \begin{subtable}{4.5cm}%
    \caption{}
      \begin{tabular}{cccc}
        \hline\hline % & \multicolumn{3}{c}{List} \\
        Rank & $R^{-1}_1$ & $R^{-1}_2$ & $R^{-1}_3$ \\ \hline
        1 & A & A & B \\
        2 & B & C & A \\
        3 & C & D & E \\
        4 & D & B & C \\
        5 & E & E & D \\ \hline
    \end{tabular}
  \end{subtable}
% \hfill
\hspace{1em}
%%%
  \begin{subtable}{4cm}%
    \caption{}
    \begin{tabular}{cccc} \hline\hline
  % & \multicolumn{3}{c}{List} \\
    Item & $R_1$ & $R_2$ & $R_3$ \\ \hline
    A & 1 & 1 & 2 \\
    B & 2 & 4 & 1 \\
    C & 3 & 2 & 4 \\
    D & 4 & 3 & 5 \\
    E & 5 & 5 & 3 \\ \hline
  \end{tabular}
\end{subtable}
% \hfill
\hspace{1em}
%%%
\begin{subtable}{4cm}%
  \caption{}
\begin{tabular}{cc} \hline\hline
% Depth   &     \\
Depth  &  $S_d$ \\ \hline
1 & $\{$A, B$\}$\\
2 & $\{$A, B, C$\}$ \\
3 & $\{$A, B, C, D, E$\}$ \\
4 & $\{$A, B, C, D, E$\}$ \\
5 & $\{$A, B, C, D, E$\}$ \\ \hline
\end{tabular}
\end{subtable}
}
\end{center}
\label{tab:example}
\end{table}

The agreement of the lists regarding the rank given to an item $X_p$ can be
measured by
\begin{equation}
  A(X_p) = f(R_1(X_p), \ldots, R_L(X_p))
\end{equation}
for a measurable non-negative real-valued distance function $f$.
For an integer $1\le d\le P$ we define the unique set of items found
by merging the first $d$ elements of each of the $L$ lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{R_l^{-1}(r) ; r \leq d, l = 1, \ldots, L \},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.
We then define the \emph{sequential rank agreement} as pooled expected agreement of the items
found in the set $S_d$:
\begin{equation}
\textrm{sra}(d)= \frac{\sum_{\{p \in S_d\}}(L-1)\E[A(X_p)]}{(L-1)|S_d|}\label{def:sra}
\end{equation}
where $|S_d|$ refers to the cardinality of the set $S_d$, and the expectation
is taken with respect to the distribution of the rankings of $X_p$ accross the $L$ lists.
Values of \textrm{sra} close to zero suggests that the lists agree while larger
values suggest disagreement. Just like in method comparison studies we
prefer smaller limits of agreement because that means that the
differences between the methods is small. If the ranked lists are
identical then the value of sequential rank agreement will be zero for
all depths $d$.

%Note that the terms that are part of \eqref{def:sra} are not
%independent since every list contains the ranks from 1 to $P$ exactly
%once. As a consequence we simply use the pooled standard deviation as
%a measure of the average distance between rankings and we refrain from
%using traditional distributional results for this measure.

Throughout the rest of this paper we will use the variance to quantify the
agreement of the rankings and therefore define
\begin{align}
  A(X_p) = (R_i(X_p) -  \E[R_i(X_p)])^2
\end{align}
For each item, the function $A$ has an interpretation as the expected
euclidean distance of the individual rankings from the expected ranking over all the lists, and
it corresponds to the same measure that is used in method comparison studies to compute the
limits of agreement \citep{alt:bland:1983}. The sequential rank agreement then becomes equivalent
to the pooled variance of the items found in $S_d$, and it can be interpreted as
the average distance of the individual rankings of the lists from the
average ranking for each of the items we have seen until depth $d$.

Given a sample of $1,\ldots,L$ lists and estimator of estimated sequatial rank agreement is given by
\begin{equation}
\widehat{\textrm{sra}}_L(d)= \frac{\sum_{\{p \in S_d\}}(L-1)\widehat{A_L}(X_p)}{(L-1)|S_d|}\label{def:sra}
\end{equation}
where
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p)
\end{align}
is the sample variance of the rank of item $X_p$. For an easier interpretation we suggest to plot $\widehat{\textrm{sra}}_L(d)^{1/2}$ corresponding to the pooled standard deviation.

\subsection{Agreement among fully observed lists}
\label{sec:amfol}

The simplest case occurs when all $L$ lists are fully observed, \ie,
we have observed the rank of all $P$ items for all $L$ lists. Fully
observed ranked lists are common and arise, for example, when
different statistical analysis methods are applied to a single dataset
to produce lists of predictors ranked according to their importance,
or if the same analysis method is applied to data from different
populations.

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to classify between acute lymphoblastic leukemia
(ALL) and acute myeloid leukemia (AML)
\citep{Golub1999}. Preprocessing of the gene expression data was done
as described in \citet{Dudoit2002} and the four different analysis
approaches were: marginal two-sample $t$ tests, marginal logistic
regression analyses, elastic net logistic regression
\citep{friedman2010regularization}, and marginal maximum information
content correlations (MIC) \citep{Reshef2011}. For the first two
methods, the genes were ranked according to minimum $p$ value, for
logistic regression the genes were ordered by size of the
corresponding coefficients (after standardization), and MIC was
ordered by absolute correlation which resulted in the top rankings
seen in Table~\ref{tab1}.

The sequential rank agreement plotted in Figure~\ref{fig:example1}
shows how many ranks apart we on average expect the genes found in the
top $d$ part of the lists to be for each depth $d$.  It can be seen
that the sequential rank agreement limits are better towards the top
of the lists (smaller values on the $y$ axis) corresponding to
\emph{better} agreement than in the tails of the lists. It is clear
from the curve in Figure~\ref{fig:example1} that there is a
substantial drop in agreement even after the first depth. Thus, if we
were to restrict attention to a small set of predictors then we would
focus on predictors 2124, 829, and 378 as seen in Table~\ref{tab1}.

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Thu Mar 31 16:50:36 2016
\begin{table}[tb]
\centering
\caption{Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method} 
\label{tab1}
\begin{tabular}{rrrrr}
  \hline
Ranking & T & LogReg & ElasticNet & MIC \\ 
  \hline
1 & 2124 & 2124 & 829 & 378 \\ 
  2 & 896 & 896 & 2124 & 829 \\ 
  3 & 2600 & 829 & 2198 & 896 \\ 
  4 & 766 & 394 & 1995 & 1037 \\ 
  5 & 829 & 766 & 808 & 2124 \\ 
  6 & 2851 & 2670 & 523 & 808 \\ 
  7 & 703 & 2939 & 1042 & 108 \\ 
  8 & 2386 & 2386 & 2489 & 515 \\ 
  9 & 2645 & 1834 & 1665 & 2670 \\ 
  10 & 2002 & 378 & 849 & 2600 \\ 
   \hline
\end{tabular}
\end{table}

\begin{figure}[htbp]
   \begin{center}
 \includegraphics[width=.49\textwidth]{paper-fig1a}
 \includegraphics[width=.49\textwidth]{paper-fig2a}
 \includegraphics[width=.49\textwidth]{paper-fig1b}
 \includegraphics[width=.49\textwidth]{paper-fig2b}
 \end{center}
 \caption{Left panels: Sequential rank agreement for 4 different
   analysis methods applied to the 3051 genes in the Golub data /black
   line). Right
   panels: Corresponding sequential rank agreement for the same data
   but where only the top 20 ranked items are available. The blue and
   red areas correspond to the independent and randomized reference
   hypothesis areas, respectively. The bottom plots are identical to
   the corresponding top plots but have been zoomed in on the first
   part of the $x$ axis.}
 \label{fig:example1}
 \end{figure}

 If the lists agree on all items then the sequential rank agreement is
 zero for all depths. Generally, if there are changes in the level of
 rank agreement then this suggests that there are sets of items that
 are ranked similarly in all lists and other sets of items that have
 been assigned vastly different ranks in the lists. In gene
 association studies we generally expect the agreement among the lists
 to be better towards the top of the lists and worse towards the end
 of the lists. In this case the sequential rank agreement curve will
 start at a low level and then increase until it levels off as in
 Figure~\ref{fig:example1}.

\subsection{Analysis of partial/censored lists}

Incomplete or partial lists are also a common occurrence. They arise
for example in case of missing data (items), when comparing top $d$
list results from publications, or when some methods only rank a
subset of the items. For example, penalized regression based on the
Lasso provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero and thus we end up with a
partial ordering. Incomplete lists also occur when the analyst decides
to censor all lists at depth $d$ such that the measurement of
agreement is restricted to the first $d$ elements of each list, or
when the analyst only wants to consider the ranking of items that have
been found to be statistically significant.

Sequential rank agreement can be generalized to partial lists in the
following way. Let $\Lambda_l\subset X$ be the subset of $d_l$ items
that have been ranked highest in list $l$.  The case where all lists are
censored at the same depth $d$ corresponds to $d_1=\cdots=d_L=d$. For
incomplete/censored lists the rank function becomes
\begin{equation}
  \widetilde R_l(X_p) = \left\{\begin{array}{cl} \{R_l(X_p)\} & \text{ for } X_p\in \Lambda_l \\
      \{d_l+1,\dots,P\} & \text{ for } X_p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

The agreement, $A(X_p)$, cannot be computed directly for all
predictors in the presence of censored lists because the exact rank
for some items will be unknown. Also, recall that the rankings within
a single list are not independent since each rank must appear exactly
once in each list. Thus, we cannot simply assign the same number
(e.g., the mean of the unassigned ranks) to the censored items since
that would result in less variation of the ranks and hence less
variation of the agreement, and it would artificially introduce a
(downward) bias of agreement for items that are censored in multiple
lists.

Instead we randomize the ranks $\{d_{l}+1,\dots,P\}$ to the items that
do not occur in list $\Lambda_l$. One realization of the $L$ rankings
of the set $X$ is obtained by randomizing the missing items of each
list. By randomizing a large number of times we can compute
\eqref{def:sra} for each realization, and then compute the sequential
rank agreement as the pointwise (for each depth) average of the rank
agreements. The algorithm is described in detail in
Algorithm~\ref{sra-algorithm}.

\begin{algorithm}
\caption{Sequential rank agreement algorithm for censored lists}
\label{sra-algorithm}
\begin{algorithmic}[1]
  \Procedure{Censored rank agreement}{} \State Let $B$ be the number
  of permutations to use \For{each $b \in B$} \For{each censored list
    $l \in L$}
  \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1.7cm}{Permute
    the unassigned ranks, $\{d_l+1, \ldots, P\}$, and assign them
    randomly to the items \emph{not} found in the list, i.e.,
    $\Lambda^\complement=X\setminus\Lambda_l$. Combine the
    result with $\Lambda_l$ to fill out the list.}
\EndFor
\State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1cm}{Let sra($b$) be the sequential rank agreement computed from the
filled out lists.}
\EndFor
\State Return element-wise averages across all $B$ permutations of sra($b$).
\EndProcedure
\end{algorithmic}
\end{algorithm}

The proposed approach is based on two assumptions: 1) that the most
interesting items are found in the top of the lists, and 2) that the
censored rankings provide so little information that it is reasonable
to assume that they can be represented by a random order. The first
assumption is justifiable because we have already accepted that it is
reasonable to rank the items in the first place. The second assumption
is fair in the light of the first assumption provided that we have a
``sufficiently large'' part of the top of the lists available.

When the two assumptions are satisfied then it is clear that the
interesting part of the sequential rank agreement curves is restricted
to the depths where the number of censored items is
low. Generally, without additional prior knowledge about the
distribution of the censored items it seems reasonable to restrict the
attention of the sequential rank agreement to depths smaller than
$\max(d_1, \ldots, d_L)$ where at least one list provides an
uncensored item.

Like for fully observed lists we generally expect the sequential rank
agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For partially ranked lists we also
expect a change-point around the depth where the lists are
censored. This is an artefact stemming from the fact that we assume
that the remainder of the lists can be replaced by a simple
permutation of the missing items.

\section{Evaluating sequential rank agreement}

To interpret the sequential rank agreement values we propose two
different benchmark values corresponding to two different
hypotheses. We wish to determine if we observe better agreement than
what would be expected if there were no relevant information available
in the data.

The first reference hypothesis is
\begin{eqnarray*}
H_0  & : &  \text{The list rankings correspond to complete randomly}\\
       &  & \text{permuted lists},
\end{eqnarray*}
which not only assumes that there is no information in the data on
which the rankings are based but also that the methods are completely
independent.

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the
method/approaches to the same data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data containing}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}
$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the fully observed case each experiment
contains $L$ lists of random permutations of the items in $X$. For the
censored/partially observed case we first permute the items
$X_1,\dots,X_P$ and then censor list $l$ at $d_l$. The sequential rank
agreement curve from the the original lists can then be compared to,
say, the pointwise 95\% quantiles of the observed rank agreements
obtained under $H_0$.

To obtain the distribution under $\widetilde H_0$ the idea is to
repeat the ranking procedures for unassociated data many times.  Thus,
we first permute the outcome variable of the data set on which the
rankings are based. This removes any association between the predictor
variables and the outcome. Then we apply the methods to the permuted
data set to generate $L$ new rankings and subsequently compute the
sequential rank agreement. Note that we only permute the outcomes and
preserve the structure of the candidate predictors. The randomization
approach requires that we have the original data available and as such
it may not be possible to evaluate $\widetilde H_0$ in all situations.

If the sequential rank agreement for the original data lies
substantially below the distribution of the sequential rank agreements
obtained under either $H_0$ or $\widetilde H_0$ then this suggests that
the original ranked lists agree \emph{more} than what we would expect
in data with no information, and therefore that the information in the
lists is significantly more in agreement than what would be expected.

Figure~\ref{fig:example1} shows the empirical distributions of
sequential rank agreement under $H_0$ and $\widetilde H_0$ each based
on $400$ permutations of the Golub data described in Section
\ref{sec:amfol}. Not surprisingly, the sequential rank agreement under
$\widetilde H_0$ is lower than the sequential rank agreement under
$H_0$ because the four methods used to rank the data ($t$ test,
logistic regression, elastic net, and MIC) generally tend to identify
(and rank) similar predictors even if there are only spurious
associations. The two bottom panels in Figure~\ref{fig:example1} also
indicate that the observed sequential rank agreement (the black line)
is better than what would be expected by chance for data that contain
no information since it lies below the reference areas. However, the
censored data also suggests that there may be at most 1 or 2 ranked
items towards the top of the lists that yield a result better than
what would be expected (the bottom-right plot).

It is important to stress that neither $H_0$ nor $\widetilde H_0$ are
related to questions regarding the association between the outcome and
the predictors in the data set. Both hypotheses are purely considering
how the rankings agree in a situation where there is no relevant
information available in the data used for creating the rankings.

\section{Change-point analysis}

In many applications it is of interest to estimate a list depths which
satisfies a change point criterion. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As Hall and Schimek, we work with an infinite set of lists
$\mathcal{L}$ and study the asymptotic behaviour for
$L\to\infty$. However, we do not allow the length of the lists to
change with $L$ as in most applications the length of the lists is
fixed [FIXME: for finite $L$ are there only finitely many forskelige
lists (aber dass stoert keinen grossen geist, siger karlsson)]. First,
we define the parameter of interest. For any given threshold value $q$
a change-point $d_0(q)$ in the list agreement is defined by
\begin{equation*}
  d_0=\textrm{arginf}_{d\in\{1,\dots,L\}}\textrm{sra}(d)>q.
\end{equation*}
The value of $q$ could for example be the agreement of randomly
permuted lists [FIXME: there is only one SRA value for infinitely many
random permutations, no skygge!].

The theorem shows consistency of the plug-in estimate of
\begin{equation*}
\hat{d_0} = \textrm{arginf}_{d\in\{1,\dots,L\}}\widehat{\textrm{sra}}(d)>q
\end{equation*}
[FIXME: $\textrm{sra}$ -> $\widehat{\textrm{sra}}$ in rest of article?]

\begin{theorem}
  $\hat{d_0}\overset{L\to\infty}{\longrightarrow}d_0$

\begin{proof}
Proof: We rewrite the square of the sequential rank agreement \eqref{def:sra} as
\begin{equation}
(\textrm{sra}(d))^2= \frac 1{\sum_{p=1}^P \sum_{l=1}^L 1\{R_l(X_p)\le d\}}\sum_{p=1}^P(1-\prod_{l=1}^L 1\{R_l(X_p)>d\}) A(X_p)^2
\end{equation}
[FIXME: Andreas details the proof with Cauchy-Schwarz.]
[FIXME: Det er lidt kedelig for saadan en fin en som Andreas: Skal theoremet og bevis ikke vaere for en general agreement metric, ikke nodvendigvis den i equation (2), og kan vi charakterisere hvad den skal opfylde?]
\end{proof}
\end{theorem}

\section{Comparison to average overlap}
